---
title: "Italian Non-Dom Tax Regime: Spatial Equilibrium Analysis"
format: html
editor: visual
Project: Fiscal Amenities and Spatial Inequality
Research Question: How did Italy's 2017 non-dom tax regime affect regional
                    housing prices and housing-burden inequality?

 Methodology: Continuous Difference-in-Differences with:
               - Regional Amenity Index (tourism, crime, coastal access)
               - Event study diagnostics
               - Pre-trend validation

 Author: Winston Lingga Ho
 Date: December 2025
---

# 1. SETUP AND DEPENDENCIES

## Load libraries

```{r}
# Load libraries
library(dplyr)
library(readr)
library(ggplot2)
library(fixest)
library(modelsummary)
library(tidyverse)
library(sf)
library(eurostat)
library(giscoR)
library(viridis)
library(kableExtra)
library(corrplot)
```

# 2. DATA LOADING AND INITIAL CLEANING

```{r}
# Load main dataset
final <- read_csv("../data/final_dataset_with_AmenityIndex.csv")

#save outputs in output folder
dir.create("../output")

# Ensure correct data types
final$territory_code <- as.character(final$territory_code)
final$Region <- as.character(final$Region)
final$Year <- as.numeric(final$Year)

cat("Dataset shape:", nrow(final), "rows x", ncol(final), "columns\n")
cat("Years covered:", min(final$Year, na.rm = TRUE), "-", 
    max(final$Year, na.rm = TRUE), "\n")
```

# 3. Amenity Index construction

The Amenity Index combines pre-2017 regional characteristics: 1. Foreign tourist nights per capita (tourism intensity) 2. Crime rate per 100,000 inhabitants (inverted) 3. Coastal dummy (binary indicator) Formula: AmenityIndex = Z_Tourism - Z_Crime + Coastal

```{r}
# 3. Time-invariant amenity index (z-scores from 2015-2019 averages)
final <- final %>%
  mutate(
    z_tourism    = as.numeric(scale(foreign_pc_avg)),
    z_crime      = as.numeric(scale(crime_avg)),
    AmenityIndex = z_tourism - z_crime + coastal_dummy,
    Post = ifelse(Year >= 2017, 1, 0)
  )


# Sanity checks
cat("\n--- Final Dataset Diagnostics ---\n")
cat("Total observations:", nrow(final), "\n")
cat("Number of regions:", n_distinct(final$territory_code), "\n")
cat("Years:", min(final$Year), "-", max(final$Year), "\n")
cat("Missing values in key variables:\n")
print(colSums(is.na(final[, c("AmenityIndex", "log_HPI_average",
                              "Observation_Income_Percentage", "Post")])))

# Summary statistics for key variables
cat("\nAmenity Index Summary:\n")
print(summary(final$AmenityIndex))


```

# 4. DESCRIPTIVE STATISTICS AND VISUALIZATIONS

```{r}
# 4a. Summary statistics table
summary_stats <- final %>%
  select(log_HPI_average, Observation_Income_Percentage, AmenityIndex,
         foreign_pc_avg, crime_avg, coastal_dummy, population) %>%
  summarise(
    across(everything(), list(
      Mean = ~mean(., na.rm = TRUE),
      SD = ~sd(., na.rm = TRUE),
      Min = ~min(., na.rm = TRUE),
      Median = ~median(., na.rm = TRUE),
      Max = ~max(., na.rm = TRUE)
    ))
  ) %>%
  pivot_longer(everything()) %>%
  separate(name, into = c("variable", "stat"), sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = stat, values_from = value)

cat("\n--- Summary Statistics ---\n")
print(summary_stats)
```

```{r}
# 4b. Correlation matrix of amenity components
corr_matrix <- final %>%
  select(foreign_pc_avg, crime_avg, coastal_dummy, AmenityIndex) %>%
  cor(use = "pairwise.complete.obs")

cat("\n--- Correlation Matrix ---\n")
print(corr_matrix)
```

```{r}
# 4c. Figure 1: Choropleth map of Amenity Index
nuts2_map <- get_eurostat_geospatial(nuts_level = "2", year = 2016)

map_data <- nuts2_map %>%
  filter(CNTR_CODE == "IT") %>%
  left_join(
    final %>% select(territory_code, AmenityIndex),
    by = c("NUTS_ID" = "territory_code")
  )

fig1_amenity_map <- ggplot(map_data) +
  geom_sf(aes(fill = AmenityIndex)) +
  scale_fill_viridis_c(option = "magma", na.value = "grey90") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
  labs(
    title = "Amenity Index Across Italian Regions (NUTS-2)",
    subtitle = "Pre-2017 tourism intensity, crime rates, and coastal access",
    fill = "Amenity\nIndex"
  )

ggsave("../output/01_amenity_map.png", fig1_amenity_map, width = 10, height = 8, dpi = 300)
print(fig1_amenity_map)
```

```{r}
# 4d. Figure 2: Distribution of Amenity Index
fig2_amenity_dist <- ggplot(final, aes(x = AmenityIndex)) +
  geom_histogram(bins = 15, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = median(AmenityIndex, na.rm = TRUE)), 
             linetype = "dashed", color = "red", size = 1) +
  theme_minimal() +
  labs(
    title = "Distribution of Regional Amenity Index",
    subtitle = "Red line indicates median (used to define high/low-amenity groups)",
    x = "Amenity Index",
    y = "Frequency"
  )

ggsave("../output/02_amenity_distribution.png", fig2_amenity_dist, width = 10, height = 6, dpi = 300)
print(fig2_amenity_dist)

```

```{r}
# 4e. Figure 3: Housing price spaghetti plot
fig3_hpi_spaghetti <- ggplot(final, aes(x = Year, y = log_HPI_average,
                                        group = Region, color = Region)) +
  geom_line(alpha = 0.6, size = 0.8) +
  geom_vline(xintercept = 2017, linetype = "dashed", color = "black", size = 1) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Housing Price Trends Across Italian Regions",
    subtitle = "Dashed line indicates 2017 non-dom tax reform",
    x = "Year",
    y = "log(Housing Price Index)"
  )

ggsave("../output/03_hpi_spaghetti.png", fig3_hpi_spaghetti, width = 12, height = 7, dpi = 300)
print(fig3_hpi_spaghetti)
```

```{r}
# 4f. Figure 4: Pre-trend comparison (high vs low amenity)
final <- final %>%
  mutate(high_amenity = AmenityIndex > median(AmenityIndex, na.rm = TRUE))
pretrend <- final %>%
  group_by(Year, high_amenity) %>%
  summarise(mean_hpi = mean(log_HPI_average, na.rm = TRUE))

fig4_pretrend <- ggplot(pretrend, aes(x = Year, y = mean_hpi, 
                                      color = high_amenity, fill = high_amenity)) +
  geom_line(size = 1.1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 2017, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("FALSE" = "#E74C3C", "TRUE" = "#3498DB"),
                     labels = c("FALSE" = "Low Amenity", "TRUE" = "High Amenity")) +
  scale_fill_manual(values = c("FALSE" = "#E74C3C", "TRUE" = "#3498DB"),
                    labels = c("FALSE" = "Low Amenity", "TRUE" = "High Amenity")) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    title = "Pre-Treatment Housing Price Trends by Amenity Level",
    subtitle = "High- and low-amenity regional trends before and after the 2017 reform",
    x = "Year",
    y = "log(Housing Price Index)",
    color = "Group",
    fill = "Group"
  )

ggsave("../output/04_pretrend_by_amenity.png", fig4_pretrend, width = 10, height = 7, dpi = 300)
print(fig4_pretrend)
```

```{r}
# 4g. Figure 5: Scatterplot of amenity vs pre-period HPI
pre_period_data <- final %>%
  filter(Year < 2017) %>%
  group_by(territory_code) %>%
  summarise(
    pre_hpi = mean(log_HPI_average, na.rm = TRUE),
    AmenityIndex = first(AmenityIndex),
    .groups = "drop"
  ) %>%
  filter(!is.na(AmenityIndex))

fig5_amenity_scatter <- ggplot(pre_period_data, aes(x = AmenityIndex, y = pre_hpi)) +
  geom_point(size = 3, color = "darkblue", alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "red", alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "Relationship Between Amenities and Pre-Reform Housing Prices",
    subtitle = "Pre-period average (2011–2016)",
    x = "Amenity Index",
    y = "log(Housing Price Index)",
    caption = "Fitted line shows simple linear relationship"
  )

ggsave("../output/05_amenity_hpi_scatter.png", fig5_amenity_scatter, width = 10, height = 7, dpi = 300)
print(fig5_amenity_scatter)

```

# 5. DIFFERENCE-IN-DIFFERENCES ANALYSIS

```{r}
# Remove observations with missing key variables for analysis
analysis_data <- final %>%
  filter(!is.na(log_HPI_average),
         !is.na(Observation_Income_Percentage),
         !is.na(AmenityIndex),
         !is.na(Post))

cat("\n--- Analysis Sample Size ---\n")
cat("Total observations:", nrow(analysis_data), "\n")
cat("Unique regions:", n_distinct(analysis_data$territory_code), "\n")
cat("Unique years:", n_distinct(analysis_data$Year), "\n")
cat("Pre-reform observations:", nrow(filter(analysis_data, Post == 0)), "\n")
cat("Post-reform observations:", nrow(filter(analysis_data, Post == 1)), "\n")
```

```{r}
# 5a. Main DiD specification: Housing Prices
did_hpi <- feols(
  log_HPI_average ~ AmenityIndex * Post | territory_code + Year,
  data = analysis_data,
  cluster = ~territory_code
)

cat("\n--- DiD Estimate: Housing Prices ---\n")
cat("Coefficient (AmenityIndex × Post):", coef(did_hpi)["AmenityIndex:Post"], "\n")
cat("Std. Error:", sqrt(diag(vcov(did_hpi)))["AmenityIndex:Post"], "\n")
summary(did_hpi)
```

```{r}
# 5b. Secondary DiD specification: Housing-Cost Burdens
did_inequality <- feols(
  Observation_Income_Percentage ~ AmenityIndex * Post | territory_code + Year,
  data = analysis_data,
  cluster = ~territory_code
)

cat("\n--- DiD Estimate: Housing-Cost Burdens ---\n")
cat("Coefficient (AmenityIndex × Post):", coef(did_inequality)["AmenityIndex:Post"], "\n")
cat("Std. Error:", sqrt(diag(vcov(did_inequality)))["AmenityIndex:Post"], "\n")
summary(did_inequality)
```

# 6. EVENT STUDY DIAGNOSTICS

```{r}
# The continuous event study validates the parallel trends assumption and
# provides year-by-year treatment effects.

analysis_data <- analysis_data %>%
  mutate(rel_year = Year - 2017)

# Event study: Housing Prices
event_hpi <- feols(
  log_HPI_average ~ 
    i(rel_year, AmenityIndex, ref = -1) | territory_code + Year,
  data = analysis_data,
  cluster = ~territory_code
)

# Extract event study coefficients for plotting
event_coefs <- broom::tidy(event_hpi) %>%
  filter(grepl("AmenityIndex", term)) %>%
  mutate(
    rel_year = as.numeric(gsub(".*rel_year::(\\-?\\d+).*", "\\1", term)),
    estimate = estimate,
    std.error = std.error,
    ci_lower = estimate - 1.96 * std.error,
    ci_upper = estimate + 1.96 * std.error
  ) %>%
  filter(!is.na(rel_year))

fig6_event_study <- ggplot(event_coefs, aes(x = rel_year, y = estimate)) +
  geom_point(size = 3, color = "darkblue") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_vline(xintercept = -0.5, linetype = "dashed", color = "black", alpha = 0.5) +
  theme_minimal() +
  theme(panel.grid.major.x = element_line(color = "grey90")) +
  labs(
    title = "Continuous Event Study: Effect of Amenity on Housing Prices",
    subtitle = "Pre-2017 coefficients should be near zero (parallel trends); dashed line at 2017 reform",
    x = "Years Relative to 2017 Reform",
    y = "Coefficient (95% CI)",
    caption = "Negative rel_year = before reform; Positive = after reform"
  )

ggsave("../output/06_event_study.png", fig6_event_study, width = 11, height = 7, dpi = 300)
print(fig6_event_study)
```

# 7. ROBUSTNESS AND DIAGNOSTIC CHECKS

```{r}
# 7a. Pre-trend regression (2011-2016 only)
pretreat_data <- analysis_data %>%
  filter(Year < 2017)

pretrend_reg <- feols(
  log_HPI_average ~ AmenityIndex * Year | territory_code + Year,
  data = pretreat_data,
  cluster = ~territory_code
)

cat("\n--- Pre-Trend Regression (2011-2016) ---\n")
cat("Testing for differential trends before reform...\n")
summary(pretrend_reg)
```

```{r}
# 7b. Balance check visualization
balance_check <- analysis_data %>%
  group_by(high_amenity, Post) %>%
  summarise(
    mean_hpi = mean(log_HPI_average, na.rm = TRUE),
    n_obs = n(),
    .groups = "drop"
  )

cat("\n--- Balance Check: Mean HPI by Group and Period ---\n")
print(balance_check)
```

# 8. EXPORT RESULTS

```{r}
# Create comprehensive results table
results_table <- modelsummary(
  list("Housing Prices" = did_hpi, 
       "Housing-Cost Burdens" = did_inequality),
  output = "gt",
  fmt = 3,
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  gof_omit = "IC|Log|Within|Pseudo|F"
)
```

```{r}
# Export as multiple formats
modelsummary(
  list("Housing Prices" = did_hpi, 
       "Housing-Cost Burdens" = did_inequality),
  output = "../output/results_table.tex",
  fmt = 3,
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  gof_omit = "IC|Log|Within|Pseudo|F"
)

modelsummary(
  list("Housing Prices" = did_hpi, 
       "Housing-Cost Burdens" = did_inequality),
  output = "../output/results_table.html",
  fmt = 3,
  stars = c('*' = 0.1, '**' = 0.05, '***' = 0.01),
  gof_omit = "IC|Log|Within|Pseudo|F"
)

cat("\n✓ Results tables exported to tables/ directory\n")

```

# 9. SESSION INFO AND SUMMARY

```{r}
cat("\n==============================================================================\n")
cat("ANALYSIS COMPLETE\n")
cat("==============================================================================\n")
cat("\nKey Findings:\n")
cat("- Housing Price Effect:", round(coef(did_hpi)["AmenityIndex:Post"], 4), "\n")
cat("- Housing Burden Effect:", round(coef(did_inequality)["AmenityIndex:Post"], 4), "\n")
cat("\nOutput files saved:\n")
cat("  Figures: 01_amenity_map.png through 06_event_study.png\n")
cat("  Tables:  results_table.tex, results_table.html\n")
cat("\nSession info:\n")
sessionInfo()
```
